---
title: "03.IAP_Structure.Rmd"
author: "Noah Kessler"
date: "13/07/2020"
output:
  html_document:
    toc: true
    toc_float: 
      collapse: false
    code_folding: "hide"
params:
  force_metasub: FALSE
---

```{r setup, message=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(Biostrings)
library(BSgenome.Mmusculus.UCSC.mm10)
library(reticulate)
use_python(Sys.which("python3"))

source("code/data_functions.R")
source("code/genomics_functions.R")
source("code/LTR_components.R")

mm10.IAP.gr <- readRDS("data/R_objects/mm10.IAP.gr.RDS")
mm10.IAP.reduced <- readRDS("data/R_objects/mm10.IAP.reduced.RDS")

meta_subelem_file <- "data/repeat_annotations/mm10.IAP.meta_subelem.tsv"
run_metasub = params$force_metasub || !file.exists(meta_subelem_file)

knitr::opts_chunk$set(message = FALSE) 
```

# Structure of IAPs

What kind of IAP element structure do we see at cVM-IAPs and tsVM-IAPs? How does this compare with the whole genome? 

** Need to run next section; maybe in new file? either way absolute paths need to be fixed, and interpretation below checked **

```{r structure.of.MEs, fig.align="center", fig.width=8, fig.height=8}
mm10.IAP.top_names <- mm10.IAP.gr %>% as.data.frame() %>%
  dplyr::select(element.ID, repName, is.LTR, width) %>%
  group_by(element.ID, is.LTR, repName) %>%
  summarize(total.bp=sum(width)) %>%
  group_by(element.ID, is.LTR) %>%
	arrange(desc(total.bp)) %>%
  filter(1:n()==1) %>% 
  ungroup()

saveRDS(mm10.IAP.top_names, "data/R_objects/mm10.IAP.top_names.RDS")

mm10.IAP.top_name.structure <- mm10.IAP.top_names %>%
	arrange(element.ID, !is.LTR) %>%
  group_by(element.ID) %>%
  summarize(top.names=paste(repName, collapse=" – "))

mm10.IAP.reduced <- left_join(mm10.IAP.reduced, mm10.IAP.top_name.structure, by="element.ID")
mm10.IAP.reduced$optimal.structure <- with(mm10.IAP.reduced,
                                      ifelse(solo.LTR | fully.structured,
                                             top.names,
                                             reduced.structure))
mm10.IAP.reduced$elem.class <- ifelse(mm10.IAP.reduced$solo.LTR, "Solo LTR", "Fully-structured or Incomplete")

mm10.IAP.optimal.counts <- mm10.IAP.reduced %>%
	group_by(optimal.structure, elem.class) %>%
	summarize(count=n())

optimal.counts.plt.df <- rbind(
	mm10.IAP.optimal.counts %>%
		mutate(val_category="Genome") %>%
		dplyr::select(val_category, everything()),
	read_tsv("data/validation/validation.element_IDs.tsv") %>%
		left_join(mm10.IAP.reduced, by="element.ID") %>%
		group_by(val_category, optimal.structure, elem.class) %>%
		summarize(count=n())
)

optimal.counts.plt.norm.df <- optimal.counts.plt.df %>%
	group_by(val_category, elem.class) %>%
	mutate(count.norm=count/max(count)) %>%
	ungroup()


category.colors <- c(Genome = "#444444", Constitutive = "#1b9e77", `Tissue-specific` = "#d95f02", `False positive` = "#cccccc")

optimal.plt.categories <- c("Genome", "Constitutive", "Tissue-specific")
optimal.plt.structs <- optimal.counts.plt.norm.df %>%
	filter(val_category %in% setdiff(optimal.plt.categories, "Genome")) %>%
	filter(count >= 1) %>%
	pull(optimal.structure) %>%
	unique

optimal.plt.st.order <- optimal.counts.plt.norm.df %>%
	filter(val_category == "Genome" & optimal.structure %in% optimal.plt.structs) %>%
	arrange(desc(count)) %>%
	pull(optimal.structure)

optimal.counts.plt.norm.df %>% 
	filter(val_category %in% optimal.plt.categories
   			 & optimal.structure %in% optimal.plt.structs) %>%
	complete(val_category,
    			 nesting(optimal.structure, elem.class),
    			 fill=list(count=0.001, count.norm=0.001)) %>%
	mutate(
		optimal.structure = forcats::fct_relabel(
			factor(optimal.structure, levels=rev(optimal.plt.st.order)),
			~ gsub("\\|", " || ", gsub("IAP-?", "", .))
		),
		val_category = factor(val_category, levels=rev(optimal.plt.categories))
	) %>%
	ggplot(aes(x=optimal.structure)) +
	  geom_bar(aes(y=count.norm, fill=val_category), stat="identity", position="dodge") +
	  coord_flip() +
		theme_bw(base_size=18) +
	  theme(panel.grid.major.y = element_blank(),
	        axis.text.y = element_text(face="bold", size=16, color = "black"),
	        axis.ticks.y = element_line(color="black", size=10),
	        axis.ticks.length.y = unit(0.03, "cm"),
	        legend.position = c(0.825, 0.075),
	        legend.background=element_rect(color="black"),
	        legend.title=element_blank(),
	        legend.margin=margin(0, 0.25, 0.25, 0.2, unit='cm')) +
	  ylab("Relative frequency of element structure") + xlab(NULL) +
	  scale_y_continuous(expand=c(0, 0.01)) +
	  scale_fill_manual(guide=guide_legend(reverse=TRUE), values=category.colors) +
	  facet_grid(vars(elem.class),  scales = "free_y", space="free_y")
```

# Internal components of IAP LTRs

Even after the mending of fragmented elements, there is still unncessary fragmentation of many LTRs and internal regions (e.g., there are frequently multiple adjacent internal sub-elements within a fully-structured element).
To better determine the size and sub-structure of the LTRs, here we combine these fragmented sub-element annotations into 'meta-subelements', so that a solo LTR only has one entry, and fully-structured IAPs only have two LTR entries and one internal entry.

```{python generate.meta.subelements, eval=run_metasub}
import sys
sys.path.insert(0, './code')
from element_metatype_boundaries import *

IAP_meta_types = load_meta_types("data/repeat_annotations/IAP_subelement_types.tsv")
IAP_elements = load_elements("data/repeat_annotations/mm10.IAP.mended.tsv", IAP_meta_types)
	
save_element_metatype_boundaries(IAP_elements, r.meta_subelem_file)
```

The distribution of the sizes of each subelement is shown below:

```{r meta.sub, fig.width=6, fig.height=2.5, fig.align='center'}
mm10.IAP.meta <- read_tsv(meta_subelem_file, col_names = c("chrom", "start", "end", "strand", "element.ID", "meta.subelement")) %>%
	separate(meta.subelement, c("subelem.idx", "subelem.type", "subelem.subidx"), remove=FALSE) %>%
	mutate(width=end-start+1)

mm10.IAP.meta %>%
	filter((subelem.type=="internal" & width < 10000) | (subelem.type=="LTR" & width < 1000)) %>%
	ggplot(aes(x=width)) +
		geom_histogram(binwidth=20, center=10) +
		theme_minimal() +
		facet_wrap(vars(subelem.type), scales="free")

saveRDS(mm10.IAP.meta, "data/R_objects/mm10.IAP.meta.RDS")
```

Summary statistics:

*LTR*

`r summary(mm10.IAP.meta[mm10.IAP.meta$subelem.type=="LTR",]$width)`

*internal*

`r summary(mm10.IAP.meta[mm10.IAP.meta$subelem.type=="internal",]$width)`

```{r find.LTR.components, fig.align='center', fig.width=6, fig.height=3}
acceptable.ltr.widths <- quantile(mm10.IAP.meta[mm10.IAP.meta$subelem.type=="LTR",]$width, c(0.025, 0.95))

LTR.gr <- mm10.IAP.meta %>%
	filter(subelem.type=="LTR" & width >= acceptable.ltr.widths[1] & width <= acceptable.ltr.widths[2]) %>%
	mutate(rowname=sprintf("%d_%s", element.ID, meta.subelement)) %>%
  tibble::column_to_rownames() %>% 
  dplyr::select(chrom, start, end, strand) %>%
	df.as.gr(use.names=TRUE)

LTR.seqs <- getSeq(BSgenome.Mmusculus.UCSC.mm10, LTR.gr)


LTR.components <- lapply(LTR.seqs, function(x) {
	tryCatch({
		suppressWarnings(identify.LTR.components(x))
	}, error = function(e) {
		message(e, "\n", x)
		return(list())
	})
})

get.R.bounds <- function(comps) c(comps$R.initial.G, comps$R.terminal.CA[2])
ltr.has.components <- unlist(lapply(LTR.components, length)) > 0

is.5p_or_solo_LTR <- endsWith(names(LTR.components), "_1:LTR:1")

R.bounds <- list()
for (ltr.name in names(LTR.components[is.5p_or_solo_LTR])) {
	ltr.comp <- LTR.components[[ltr.name]]
	if (length(ltr.comp) == 1) R.bounds[[ltr.name]] <- get.R.bounds(ltr.comp[[1]])
	else if (length(ltr.comp) > 1) {
		unique.Rs <- unique(lapply(ltr.comp, get.R.bounds))
		if (length(unique.Rs) == 1) R.bounds[[ltr.name]] <- unique.Rs[[1]]
	}
}

R.bound.frac <- lapply(names(R.bounds), function(nm) R.bounds[[nm]] / width(LTR.seqs[nm]))
bound.frac.U3_R <- unlist(purrr::map(R.bound.frac, 1))
bound.frac.R_U5 <- unlist(purrr::map(R.bound.frac, 2))

gradient.plot <- function(middles, edges, colors, height=1) {
	stopifnot(length(middles) == length(edges))
	stopifnot(all(sapply(edges, length) == 2))
	stopifnot(length(colors) == length(middles)+1)
	
	create.gradient <- function(mincolor, maxcolor, N=15, trim.end=c(FALSE, FALSE)) {
		colors <- colorRampPalette(c(mincolor, maxcolor))(N)
		if (trim.end[1]) colors <- colors[-1]
		if (trim.end[2]) colors <- colors[-N]
		grid::rasterGrob(
			t(colors),
			width = unit(1,"npc"), height = unit(1,"npc"),
			just="center", interpolate = TRUE
		)
	}
	
	plt <- ggplot(data.frame(x=c(0, 1), y=c(0, height))) +
		#geom_point(aes(x=x, y=y))
		#xlim(0, 1) + ylim(0, height)
		scale_x_continuous(limits=c(0, 1), expand=c(0, 0)) +
		scale_y_continuous(limits=c(0, height), expand=c(0, 0))
	
	for (idx in 1:length(middles)) {
		gL <- create.gradient("white", colors[idx], trim.end=c(TRUE, FALSE))
		gR <- create.gradient(colors[idx+1], "white", trim.end=c(FALSE, TRUE))
		
		m <- middles[idx]
		e <- edges[[idx]]
		
		plt <- plt + 
			annotation_custom(gL, xmin=e[1], xmax=m, ymin=0, ymax=height) +
			annotation_custom(gR, xmin=m, xmax=e[2], ymin=0, ymax=height)
	}
	
	plt +
		theme(
			axis.title = element_blank(),
			panel.grid = element_blank(),
			panel.background = element_rect(fill="white"),
			panel.border = element_rect(color="black", fill=NA, size=1),
			plot.margin = unit(c(0.03, 0.03, 0.03, 0.03), "npc"), 
			axis.text.y = element_blank(), 
			axis.ticks.y = element_blank(),
			axis.text.x = element_text(color="black"),
			axis.ticks.x = element_line(color="black")
		)
}

comp.mids <- c(
	median(bound.frac.U3_R, na.rm=TRUE),
	median(bound.frac.R_U5, na.rm=TRUE)
)
plt.quantiles <- c(0.1, 0.9)
comp.edges <- list(
	quantile(bound.frac.U3_R, plt.quantiles, na.rm=TRUE),
	quantile(bound.frac.R_U5, plt.quantiles, na.rm=TRUE)
)
comp.colors <- c("#1b9e77", "#d95f02", "#7570b3")

comp.centres <- c(mean(c(0, comp.mids[1])), mean(c(comp.mids)), mean(c(comp.mids[2], 1)))

gradient.plot(comp.mids, comp.edges, comp.colors) +
	annotate("text", x=comp.centres, y=0.5, label=c("U3", "R", "U5"), size=8)
```